---
title: "Homework 2 - data wrangling with the tidyverse"
author: "Chiamaka Azodo"
date: "2/14/2022"
output: 
  md_document:
    variant: markdown_github
    toc: yes    
  html_document:
    
    toc: yes
    toc_float: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```

```{r, include=FALSE}

library(ggplot2)
library(tidyverse)
library(janitor)
library(dplyr)

```

```{r, include=FALSE}

# Upload the data

setwd("/Users/chiamakaazodo/Documents/MDS_2022/github/assignments/homework-2-chiamakaazodo/data/")
seasons_stats <- read_csv("Seasons_Stats_NBA.csv")

```

***

### 1. Which NBA player scored the most points in 1991?

Ricky Pierce scored the most points in 1991.

```{r}

# Select the relevant columns, filter for the year of interest, then group by players so that when
# the sum of points is collected its is by player and then sort in descending order the total number of points scored 
# by the player and limit to the top 10 rows.

mvp_player <- seasons_stats %>%
  select(Year, Player, PTS) %>%
  filter(Year == 1991)  %>%
  group_by(Player) %>%
  summarise(total_pts = sum(PTS)) %>%
  arrange(desc(total_pts)) %>%


mvp_player %>% top_n(10)

```

### 2. Which player had the best free throw percentage from the year 2000 to the most recent year in the data?

Since the year 2000, Aaron Miles has had the best free throw percentage. (When sorted alphabetically due to tied percentages)

```{r}

# Select the relevant columns, filter to the year 2000 and after 
# Order the results by free throw percentage and then by player

fthrowperplayer <- seasons_stats %>%
  select(Year, Player, `FT%`) %>% 
  filter(Year >= 2000) %>%
  arrange(desc(`FT%`)) %>% 
  arrange((Player)) %>%
  

fthrowperplayer %>% top_n(10)

```

### 3. Rename the variable "Pos" to "position".

```{r}

# Use the rename function to change Pos variable. 

seasons_stats %>%
  rename(position = Pos) %>%
  top_n(10)
  
```


### 4. Use this variable to create two variables that are called "first_position" and "second_position".

```{r, warning=FALSE}

# Use the separate function to split position into two variables. 
# There are NAs recorded for some players in the second_position variable because they didn't have a hyphenated position

seasons_stats <- seasons_stats %>%
  rename(position = Pos) %>%
  separate(
    position, 
    into = c("first_position", "second_position")) 

# Preview the edited data

seasons_stats %>%
  top_n(10)

```


### 5. Unite these two variables back into a single variable called "position_united".

```{r}

seasons_stats <- seasons_stats %>%
  unite(
    position_united, 
    first_position, 
    second_position, 
    sep = "-", 
    na.rm = TRUE) 


# Preview the edited data

seasons_stats %>%
  top_n(10)

```


### 6. Create two new datasets.

```{r}

# Create a new data set with all variables excluding age using the select function
# Also use select to create a new data with just player information
# Use the mutate and seq functions to create a new column that counts from 1, to the length of the data table, by 1 intervals
# Move mergeid to the front of the table using relocate for easier visibility

season_data <- seasons_stats %>%
  select(-Age) %>%
  mutate(mergeid = seq(1, nrow(seasons_stats), 1)) %>%
  relocate(mergeid)

player_data <- seasons_stats %>%
  select(
    Year, 
    Player, 
    Age) %>%
  mutate(mergeid = seq(1, nrow(seasons_stats), 1)) %>%
  relocate(mergeid)

# Preview the completed data sets

season_data %>%
  top_n(10)

player_data %>%
  top_n(10)
```


### 7. Join the two datasets from question (6) together to recreate the original dataset plus the new merge id.

```{r}

# Use an inner join to combine the two tables since all records should appear in both

merged_data <- player_data %>%
  inner_join(
    season_data, 
    by = "mergeid")

# Preview the merged data

merged_data %>%
  top_n(10)

```


### 8. Subset the original dataset to 1995. Group the data by year and team name and then summarize the average number of points per team. Arrange from most to least points.

```{r, message=FALSE}

# Summarizing points scored per team for the year 1995 using dplyr functions

team_pts_95 <- seasons_stats %>%
  filter(Year == 1995) %>%
  group_by(
    Year, 
    Tm) %>%
  summarise(avg_pts = mean(PTS, na.rm = TRUE)) %>%
  arrange(desc(avg_pts)) 


team_pts_95 %>%
  top_n(10)

```

### 9. Reshape the data in the previous question into a wide format using the `tidyr` package. Create a wide dataset that keeps year in a single column, but spreads team names to multiple individual columns with each column delineating points per team in 1995.

```{r}

# Use the pivot_wider function to spread the data specifying the columns to collect the names and values from

team_pts_95_wide <- team_pts_95 %>%
  pivot_wider(
    names_from = Tm, 
    values_from = avg_pts)

team_pts_95_wide %>%
  top_n(10)
```


### 10. Now return the data to a long (tidy) format by moving teams back into a single column and points in a single column.

```{r}

# Use the pivot_longer function to return the data back to normal
# Instead of listing all the column names, just exclude the year column from pivot_longer

team_pts_95_wide %>%
  pivot_longer(
    !Year, 
    names_to = "team",
    values_to = "avg_pts") %>%
  top_n(10)

```





